The effect of climate forcing on numerical simulations of the Cordilleran ice sheet at the Last Glacial Maximum

### 2014-04-03 cordillera-climate.tex
Add comments from Arjen in manuscript

% Affiliations
\affil[1]{Department of Physical Geography and Quaternary Geology, Stockholm University, Stockholm, Sweden}
\affil[2]{Geophysical Institute, University of Alaska Fairbanks, Fairbanks, AK, USA}
\affil[3]{Helmholtz Centre Potsdam, GFZ German Research Centre for Geosciences, Potsdam, Germany}

@@

At the Last Glacial Maximum (LGM), glaciers of a~size comparable to the present Greenland and Antarctic ice sheets covered parts of Northern America (Laurentide, Cordilleran and Innuitian ice sheets) and northern Eurasia (Fennoscandian ice sheet). Numerical modelling of these former ice masses allows for a~comparison between glaciological theories embedded in the models and geomorphological traces underpinning palaeo-glaciological reconstructions. Yet, a~major obstacle in this exercise resides in large uncertainties concerning the climate forcing, typically atmospheric temperature and precipitation, required as input to numerical glacier models \citep{hebeler-etal-2008}. This includes uncertainty in the representation of Earth's present climate in regions of poor station coverage, and even larger uncertainty concerning accurate reconstructions of past climate change.

@@

Because much of the disparity between reanalytic and observational temperature data is caused by topographical detail at scales unresolved in the reanalyses, we apply a lapse-rate correction prior to this comparison, using the same lapse rate of 6\,\unit{{\degree}C\,km^{-1}} as in the simulations. ERA-Interim, CFSR and NARR climatologies exhibit temperature distributions fairly consistent with WorldClim data (Fig.~\ref{fig:tempheatmap}). The spatial distribution of temperature differences between reanalyses and WorldClim data shows partly consistent patterns, such as positive anomalies in the northernmost part of the modelling domain. These may relate to weaknesses of WorldClim data in areas of sparse observational coverage (Fig.~\ref{fig:tempdiff}). However, significant disparity among reanalyses exists, too. More specifically, surface air temperature data from the NCEP/NCAR climatology shows the largest disparity with that of WorldClim, with a~significant cold anomaly over most of the modelling domain (Figs.~\ref{fig:tempheatmap} and \ref{fig:tempdiff}).

@@

Most likely, this is the signature of an orographic precipitation effect. As described in Sect.~\ref{sec:climate}, the topography of the Northern American Cordillera is such that its western ranges form a~continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid. However the ability of a~GCM to reproduce these contrasts in precipitation is bounded by spatial resolution. In a~model of coarser resolution, these high mountain ranges are reduced to smooth hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}). However, this effect alone cannot explain the widespread and pronounced positive precipitation anomalies observed in CFSR data (Figs.~\ref{fig:precheatmap} and \ref{fig:precdiff}) despite its high spatial resolution (Table~\ref{tab:reanalyses}).

@@

Although only temperature or precipitation changes are applied, this experiment results once more in large differences in modelled ice sheet geometries (Fig.~\ref{fig:biatm}) as compared to the reference WorldClim -5\,\unit{{\degree}C} run (Fig.~\ref{fig:cool05}, top-left panel). For ERA-Interim, CFSR and NARR climatologies, precipitation anomalies clearly dominate the differences in ice sheet response, whereas temperature anomalies have relatively little effect (Fig.~\ref{fig:biatmbars}). However in the case of NCAR forcing, both the negative temperature bias (Figs.~\ref{fig:tempheatmap} and \ref{fig:tempdiff}) and the important precipitation differences (Figs.~\ref{fig:precheatmap} and \ref{fig:precdiff}) contribute approximately equally to produce oversized ice sheets (Fig.~\ref{fig:biatmbars}).

\subsection{Comparison to geomorphological LGM margin}

@@

This experiment shows that a~shorter start-up period leads to more restricted ice cover in the eastern part of the modelling domain, where precipitation rate controls ice advance, but more extensive glaciation in the south-western domain along the Pacific coast, where temperature is the limiting factor (Fig.~\ref{fig:durationstack}). Hence, reducing simulation duration leads to a~closer match with the LGM ice sheet margin by \citet{dyke-2004} in some parts of the modelling domain, but higher discrepancy in other parts.

>>>>

% Affiliations
\affil[1]{Department of Physical Geography and Quaternary Geology and the Bolin Centre for Climate Research, Stockholm University, Stockholm, Sweden}
\affil[2]{Geophysical Institute, University of Alaska Fairbanks, Fairbanks, AK, USA}
\affil[3]{Helmholtz Centre Potsdam, GFZ German Research Centre for Geosciences, Potsdam, Germany}

@@

At the Last Glacial Maximum (LGM), glaciers of a~size comparable to the present Greenland and Antarctic ice sheets covered parts of Northern America (Laurentide, Cordilleran and Innuitian ice sheets) and northern Eurasia (Fennoscandian and Eurasian ice sheets). Numerical modelling of these former ice masses allows for a~comparison between glaciological theories embedded in the models and geomorphological traces underpinning palaeo-glaciological reconstructions. Yet, a~major obstacle in this exercise resides in large uncertainties concerning the climate forcing, typically atmospheric temperature and precipitation, required as input to numerical glacier models \citep{hebeler-etal-2008}. This includes uncertainty in the representation of Earth's present climate in regions of poor station coverage, and even larger uncertainty concerning accurate reconstructions of past climate change.

@@

Because much of the disparity between reanalyses and observational temperature data is caused by topographical detail at scales unresolved in the reanalyses, we apply a lapse-rate correction prior to this comparison, using the same lapse rate of 6\,\unit{{\degree}C\,km^{-1}} as in the simulations. ERA-Interim, CFSR and NARR climatologies exhibit temperature distributions fairly consistent with WorldClim data (Fig.~\ref{fig:tempheatmap}). The spatial distribution of temperature differences between reanalyses and WorldClim data shows partly consistent patterns, such as positive anomalies in the northernmost part of the modelling domain. These may relate to weaknesses of WorldClim data in areas of sparse observational coverage (Fig.~\ref{fig:tempdiff}). However, significant disparity among reanalyses exists, too. More specifically, surface air temperature data from the NCEP/NCAR climatology shows the largest disparity with that of WorldClim, with a~significant cold anomaly over most of the modelling domain (Figs.~\ref{fig:tempheatmap} and \ref{fig:tempdiff}).

@@

Most likely, this is the signature of an orographic precipitation effect. The topography of the Northern American Cordillera is such that its western ranges form a~continuous orographic barrier, causing high precipitation along the Pacific coast while leaving much of the interior arid. The ability of a~GCM to reproduce these contrasts in precipitation is bounded by spatial resolution. In a~model of coarser resolution, these high mountain ranges are reduced to smooth hills and the modelled distribution of precipitation is also smoother (Fig.~\ref{fig:oroprecip}). However, this effect alone cannot explain the widespread and pronounced positive precipitation anomalies observed in CFSR data (Figs.~\ref{fig:precheatmap} and \ref{fig:precdiff}) despite its high spatial resolution (Table~\ref{tab:reanalyses}).

@@

Although only temperature or precipitation changes are applied, this experiment results once more in large differences in modelled ice sheet geometries (Fig.~\ref{fig:biatm}) as compared to the reference WorldClim -5\,\unit{{\degree}C} run (Fig.~\ref{fig:cool05}, top-left panel). For ERA-Interim, CFSR and NARR climatologies, precipitation anomalies clearly dominate the differences in ice sheet response, whereas temperature anomalies have relatively little effect (Fig.~\ref{fig:biatmbars}). However, in the case of NCAR forcing, both the negative temperature bias (Figs.~\ref{fig:tempheatmap} and \ref{fig:tempdiff}) and the important precipitation differences (Figs.~\ref{fig:precheatmap} and \ref{fig:precdiff}) contribute approximately equally to produce oversized ice sheets (Fig.~\ref{fig:biatmbars}).

\subsection{Comparison to geomorphological LGM margin reconstruction}

@@

This experiment shows that a~shorter start-up period leads to more restricted ice cover in the eastern part of the modelling domain, where precipitation rate controls ice advance, but more extensive glaciation in the south-western domain along the Pacific coast, where temperature is the limiting factor (Fig.~\ref{fig:durationstack}). Hence, reducing simulation duration leads to a~closer match with the LGM ice sheet margin by \citet{dyke-2004} in some sectors of the modelling domain, but higher discrepancy in other sectors.
